Colonies - Compute Continuums across Platforms

### 2023-03-31 paper_2023/paper.tex
wip

\usepackage{arxiv}
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{cleveref}       % smart cross-referencing
\usepackage{lipsum}         % Can be removed after putting your text content
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{doi}
\usepackage{listings}

@@

\author{{\hspace{1mm}Johan Kristiansson} \\
	Department of Computer Science \\
	RISE Research Institutes of Sweden \\
	Luleå, Sweden \\
	\texttt{johan.kristiansson@ri.se} \\
	\And
	{\hspace{1mm}Thomas Ohlson Timoudas} \\
	Department of Computer Science \\
	RISE Research Institutes of Sweden \\
	Luleå, Sweden \\
	\texttt{thomas.ohlson.timoudas@ri.se} \\
	\And
	{\hspace{1mm}Henrik Forsgren} \\
	Department of Computer Science \\
	RISE Research Institutes of Sweden \\
	Luleå, Sweden \\
	\texttt{thomas.ohlson.timoudas@ri.se} \\
	\And
	{\hspace{1mm}Erik Källman} \\
	Department of Computer Science \\
	RISE Research Institutes of Sweden \\
	Luleå, Sweden \\
	\texttt{thomas.ohlson.timoudas@ri.se} \\
}

@@

\begin{abstract}
Artificial intelligence and machine learning has gained significant traction in recent years. At the same time, development and operation of AI workloads has become increasingly challenging. One difficulty is the lack of portability, making it cumbersome to move from one platform or provider to another. Creating and operating fully automated end-to-end workflows across devices, edge, and cloud platforms is even more challenging.

To address these issues, the paper presents a novel framework called Colonies for running computational workloads across heterogeneous platforms. Colonies is built upon a loosely coupled microservice architecture that breaks down complex workflows into composable functions, which can be executed by independently deployed executors. With the use of an HTTP protocol, the functions can be composed into declarative workflows in any computer language, and be executed across platforms by executors running in the cloud, edge, devices, or even in web browsers, creating seamless compute continuums across platforms.

@@

Taking an AI model into production requires substantial software engineering expertise. In contrast to traditional IT workloads, both the data and the model must be managed in addition to the software itself. As most models require regular re-training or re-calibration, it must be possible to update deployed models and software seamlessly without losing information or breaking the system. In many cases, there is a constant flow of data ingested into the system which must be managed even if components are malfunctioning. This becomes even more challenging when nodes or parts of the underlying infrastructure become unavailable due to maintenance or failaure such as software updates, hardware replacements or simply missconfiguration.

@@

Additionally, many compute clusters operate on-premises installations. Sometimes it is necessary to temporarily increase the capacity of on-prem clusters by combining resources from multiple providers, for example, adding cloud compute resources to handle peak loads or utilize HPC resources to quickly reprocess historical data. Developing hybrid workflows where some jobs run in the cloud and others run on HPC systems requires even more software development efforts and is beyond the scope of many users, preventing them from utilizing powerful hardware. Clearly, there is a need for a framework that can consolidate various platforms to simplify development and enable seamless execution across platforms.

This paper presents a framework called Colonies, which is built around a loosely coupled microservices architecture that separates workflow definitions from implementation and deployment. The main objective is to create a tool where monolithic workflows can be broken down into independent and separated compute units that can be dynamically added or removed while executing workflows. These compute units can be implemented in any computer language and be deployed anywhere on the Internet. Ultimately, the Colonies framework facilitates coordination among diverse software systems operating on different platforms in executing unified workflows.

The next section presents related work. Section ....

\section{Related work}
A workflow is a series of tasks that need to be completed in a specific order. Workflows are often represented as directed acyclic graphs (DAGs), where nodes represent tasks and edges represent dependencies or data flow between tasks.

@@

\section{The Colonies framework}
\label{sec:headings}
Microservices is an architectural design pattern in which an application is structured as a collection of small, independently deployable, and loosely coupled services that communicate with other microservices through well-defined APIs. By dividing the application into smaller, focused microservices, applications become easier to understand, maintain, and develop. Each microservice can be scaled independently, making it easier to handle increased demand for a certain service. Additionally, different microservices can be developed by using diverse technologies, frameworks, or programming languages, enabling developers to select the most suitable tools for each specific problem. By assigning ownership of specific microservices to individual teams, it also becomes easier to coordinate work and maintain a consistent development process.

@@

Colonies is based on a microservice architecture where workflows are decomposed into a set of functions. An example of a function could for example be to train a neural network, prepare a batch of data, or upload inferred results to a third-party database. An executor is responsible for executing one or several functions, making it very simular to a microservice. All coordination is managed by a cluster of Colonies servers, allowing complex workflows to be broken down into independent functions that can be developed and tested separately. Analogous to traditional microservices, scalability can be achieved simply by adding more executors. If an executor crashes during task execution, the task is automatically reassigned to another executor.

\subsection{The role of queues for seperation of concerns}
Separation of concerns (SoC) is a fundamental design principle in computer science that aims to break down a complex software system into smaller, more manageable parts. For example, HTTP APIs can be used to abstract away implementation detail and provide a clear and simple interface for interacting with a particular service. However, HTTP protocols alone are insufficient for handling dynamic environments where components frequently fails or the underlying infrastructures is constantly changing.

To address such environments, an alternative mechanism is necessary. Queues enable different executors to communicate indirectly by acting as a buffer between them. Queues makes it possible to decouple each executor to operate independently, e.g. a service can be updated or replaced without affecting other executors. Queues also allow for asynchronous communication between executors, enabling them to process tasks at their own pace. This ensures that slower executors do not bottleneck faster ones, leading to a more efficient and scalable system. Most importantely, queues enable load balancing by distributing tasks among multiple executors, thus making it possible to parallelize workflow execution.

Queues can be implemented in various ways. As previously mentioned, a common solution involves using a message broker. However, Colonies adopts an alternative approach by utilizing a standard database and querying it to assign tasks to different executors. The advantage of employing a database lies in the fine-grained control it offers for assigning task to executors. For instance, it allows a specific tasks be assigned to particular executor type, thus enabling a workflow to be executed across diverse platforms.

\begin{figure}[h]
	\centering
    \includegraphics[scale=0.4]{arch.png}
	\caption{cron management}
	\label{fig:fig1}
\end{figure}

\begin{figure}[h]
	\centering
    \includegraphics[scale=0.4]{raft.png}
	\caption{cron management}
	\label{fig:fig1}
\end{figure}

\subsubsection{Workflows}
TODO
\begin{figure}[h]
	\centering
    \includegraphics[scale=0.32]{workflow.png}
	\caption{cron management}
	\label{fig:fig1}
\end{figure}

>>>>

\usepackage{arxiv}
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{cleveref}       % smart cross-referencing
\usepackage{lipsum}         % Can be removed after putting your text content
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{doi}
\usepackage{listings}
\usepackage{caption}
\usepackage{subcaption}

@@

\author{{\hspace{1mm}Johan Kristiansson} \\
	Department of Computer Science \\
	RISE Research Institutes of Sweden \\
	Luleå, Sweden \\
	\texttt{johan.kristiansson@ri.se} \\
	\And
	{\hspace{1mm}Thomas Ohlson Timoudas} \\
	Department of Computer Science \\
	RISE Research Institutes of Sweden \\
	Luleå, Sweden \\
	\texttt{thomas.ohlson.timoudas@ri.se} \\
	\And
	{\hspace{1mm}Henrik Forsgren} \\
	Department of Computer Science \\
	RISE Research Institutes of Sweden \\
	Luleå, Sweden \\
	\texttt{thomas.ohlson.timoudas@ri.se} \\
	\And
	{\hspace{1mm}Erik Källman} \\
	Department of Computer Science \\
	RISE Research Institutes of Sweden \\
	Luleå, Sweden \\
	\texttt{erik.kallman@ri.se} \\
}

@@

\begin{abstract}
Artificial intelligence and machine learning has gained significant traction in recent years. At the same time, development and operation of AI workloads has become increasingly challenging. One difficulty is the lack of portability, making it cumbersome to move from one platform to another. Creating and operating fully automated end-to-end workflows across devices, edge, and cloud platforms is even more challenging.

To address the aforementioned challenges, the paper introduces a framework termed Colonies designed to facilitate the execution of computational workloads across a diverse range of platforms. Colonies is founded upon a loosely-coupled microservice architecture that beaks down complex workflows into composable functions. With the use of an HTTP protocol, these composable functions can be implemented in any computer language and be executed by independent executors deployed across various systems, e.g. cloud, edge, devices, or even in web browsers. Colonies orchestrates the execution and by using a zereo-trust security protocol, a collection of distributed executors can function as a single cohesive unit, thereby establishing a seamless compute continuum across multiple platforms.

@@

Taking an AI model into production requires substantial software engineering expertise. In contrast to traditional IT workloads, both the data and the model must be managed in addition to the software itself. As most models require regular re-training or re-calibration, it must be possible to update deployed models and software seamlessly without losing information or breaking the system. In many cases, there is a constant flow of data ingested into the system which must be managed even if some components are malfunctioning. This becomes even more challenging when nodes or parts of the underlying infrastructure become unavailable due to maintenance or failaure such as software updates, hardware replacements or simply missconfiguration.

@@

Additionally, many compute clusters operate on-premises installations. Sometimes it is necessary to temporarily increase the capacity of on-prem clusters by combining resources from multiple providers, for example, adding cloud compute resources to handle peak loads or utilize HPC resources to quickly reprocess historical data. Developing hybrid workflows where some jobs run in the cloud and others run on HPC systems requires even more software development efforts and is beyond the scope of most users, preventing them from utilizing powerful hardware. Clearly, there is a need for a framework that can consolidate various platforms to simplify development and enable seamless execution across platforms.

This paper presents a framework called Colonies, which is built around a loosely-coupled microservices architecture that separates workflow definitions from implementation and deployment. The main objective is to create a tool where monolithic workflows can be broken down into independent and separated compute units that can be dynamically added or removed while executing workflows. These compute units can be implemented in any computer language and be deployed anywhere on the Internet.

The next section presents related work. The next section describes related work. Section X ...

\section{Related work}
A workflow is a series of tasks that need to be completed in a specific order. Workflows are often represented as directed acyclic graphs (DAGs), where nodes represent tasks and edges represent dependencies or data flow between tasks. Workflow management has been extensively studied in both academic and industrial settings, with numerous approaches proposed to address the challenges in this field.

Another popular approach is the use of workflow engines, which provide a framework for creating and executing workflows. Workflow engines can be integrated with other tools and systems to automate complex business processes. Examples of workflow engines include Apache Airflow, Camunda, and Zeebe.

Apache Airflow is a popular open-source workflow management system used to programmatically author, schedule, and monitor workflows.
Luigi is also an open-source workflow management system developed by Spotify. It is designed to handle long-running batch jobs and data pipelines.

Pinball is a workflow management system developed by Pinterest that allows users to define and execute workflows using Python scripts. It has a simple architecture, with support for DAGs and task dependencies. However, it lacks some of the advanced features of Airflow, such as task retries and monitoring.

@@

\section{The Colonies framework}
\label{sec:headings}
Microservices is an architectural design pattern in which an application is structured as a collection of small, independently deployable, and loosely coupled services that communicate with other microservices through a well-defined API. By dividing the application into smaller, focused microservices, applications become easier to understand, maintain, and develop. Each microservice can be scaled independently, making it easier to handle increased demand for a certain service. Additionally, different microservices can be developed by using diverse technologies, frameworks, or programming languages, enabling developers to select the most suitable tools for each specific problem. By assigning ownership of specific microservices to individual teams, it also becomes easier to coordinate work and maintain a consistent development process.

@@

Colonies is based on a microservice architecture where workflows are decomposed into a set of independent functions. An example of a function could for example be a function that trains a neural network, prepare a batch of data, or upload inferred results to a third-party database etc. An executor is responsible for executing one or several functions, making it very simular to a microservice. All coordination is managed by a cluster of Colonies servers, allowing complex workflows to be broken down into independent functions that can be developed and tested separately. Analogous to traditional microservices, scalability can be achieved simply by adding more executors implementing the same function specification. If an executor crashes during task execution, the task is automatically reassigned to another executor.

\subsection{The role of queues as seperation of concerns}
Separation of concerns (SoC) is a fundamental design principle in computer science that aims to break down a complex software system into smaller, more manageable parts. For example, HTTP APIs can be used to abstract away implementation detail and provide a clear and simple interface for interacting with a particular service. However, HTTP protocols alone are insufficient for handling dynamic environments where components frequently fails or the underlying infrastructures is constantly changing. To address such environments, an alternative mechanism is necessary.

Queues enable different software components to communicate indirectly by acting as a buffer between them. Queues makes it possible to decouple each executor and make them operate independently, e.g. an executor can be updated or replaced without affecting other executors. Queues also allow for asynchronous communication between executors, enabling them to process tasks at their own pace. This ensures that slower executors do not bottleneck faster ones, leading to a more efficient and scalable system. Most importantely, queues enable load balancing by distributing tasks among multiple executors, thus making it possible to parallelize workflow execution.

Queues can be implemented in different ways, and while message brokers are a common solution, Colonies has adopted an alternative strategy by leveraging a standard database and querying it for tasks to assign to different executors. One key advantage of this approach is that it enables fine-grained task assignments, making it possible to assign specific tasks to particular executors. For instance, an executor of the browser type can be limited to only executing tasks of the browser type. This level of granularity cannot easily be implemented using message brokers which generally do not offer introspection of queue, or provide the ability to pull specific messages out of the queue. Generarally, the only way to retrieve a specific message is to pull all messages from the queue, obtain the message, and then place all remaining messages back into the queue in the same order. In contrast, using a database a query can match any columns and be ordered using a timestamp to make it function as a queue.

\begin{figure}
     \centering
     \begin{subfigure}[b]{0.3\textwidth}
         \centering
         \includegraphics[scale=0.39]{arch.png}
         \caption{Process assignment steps.}
         \label{fig:y equals x}
     \end{subfigure}
     \hfill
     \begin{subfigure}[b]{0.3\textwidth}
         \centering
         \includegraphics[scale=0.39]{process.png}
         \caption{Manipulation of a process.}
         \label{fig:three sin x}
     \end{subfigure}
     \caption{Process management via Colonies HTTP API. Note that only the assigned executor has write access to the process database entry.}
     \label{fig:three graphs}
\end{figure}

\subsection{Process tables}
Colonies provides a platform for executors to interact with each other by enabling them to submit function specifications to a Colonies server. Once submitted, other executors can connect to the server to receive execution assignments. When a function specification is submitted to the Colonies server, it is store as a process entry in a dedicated process table database. In this case, a process is analogous to a task, but contains additional contextual information about the execution, for example, execution status (waiting, running, successful, failed), assigned executor Id, submission time, execution time, deadlines, priority etc. It also contains input and argument values as well as return values from the function innovation.

The Colonies server acts as a job broker for executors, almost like an employment agency. When an executor connects to the Colonies server, the server hang the incomming HTTP connection\footnote{An alternative protocol is to use WebSockets or gPRC to communicate with the Colonies server.} until the executor is assigned a process, or until a connection timer expires. Note that the Colonies server does not connect to any executor. Rather, it is the responsibility of the executors to connect to the Colonies server, enabling them to be deployed anywhere on the Internet, even behind firewalls or commercial telco networks.

Figure \ref{fig:fig1} illustrates an executor submitting a function specification that is later assigned to another executor. Executors have to specify to the Colonies server which functions they are capable of executing. The Colonies server makes sure that the conditions (or requirmenets) of a function specification matches the capability of the executors.

\begin{figure}[h]
	\centering
    \includegraphics[scale=0.35]{function_spec.png}
	\caption{Example of a function spec.}
	\label{fig:fig2}
\end{figure}

Figure \ref{fig:fig2} shows an example of a function specification. The \emph{maxexectime} attribute specifies the maximum time an executor may run a process (in this case, 100 seconds). Before a process is assigned to an executor, the Colonies server updates the process entry in the process table database and calculates a deadline for when the process must be finished. The server then regularly checks for any running processes that have exceeded their deadlines. If such process is detected, it is reset, allowing it to be re-assigned to other executors. This scheme is completely stateless, so if the Colonies server crashes, another Colonies server can take over and look for processes that have exceeded their deadline. Moreover, if an executor assigned to a process crashes, the process is simply re-executed by another executor.

To scale up a system, more executors can be deployed. Scaling down, however, can be more challenging. One solution is to select a set of executors to be removed and then starv them out by denying them new process assignments. Another, simpler solution, is to immediately destroy the executors, or if possible notify them in advance so that they can gracefully unregister.

The \emph{maxexectime} failsafe mechanism ensures that processes will eventually be executed even in the case of failures. This mechanism also relieves the burden of user to check if a process has been executed or not, as they can simply look up the process in the database to get its current status. It is also possible to subscribe on process changhes using an event-driven protocol, but this may be impractical if a process runs for a very long time (e.g. days or weeks).

\begin{figure}[h]
	\centering
    \includegraphics[scale=0.4]{raft.png}
	\caption{cron management}
	\label{fig:fig2}
\end{figure}

\subsubsection{Concurreny and synchronization}
Synchronization is esstential to prevent data inconsistency and race conditions when accessing shared resources concurrently with multiple threads. However, synchronization also slows down execution as only one thread can access critical sections at a time. By carefully designing multithreaded applications and employing the right synchronization techniques, it possible to minimize the performance impact while still ensuring data consistency and correctness.

The purpose of all Colonies APIs is to either modify the database or read information from the database. A request

Colonies is designed to be stateless, meaning that the Colonies server does not keep any information about between API requests. In other words, each request is handled independently, without relying on any information from previous requests. Once the database has been updated,

if an instance fails or becomes unavailable, subsequent requests can be easily redirected to another

. After a The Colonies server ensure that only the assigned executor has write access to the process entry in the database.

Colonies ensures that exactly one executor is assigned to a specific process, so only the assigned executor has permission to set output values and close the process as successful. However, there is a window of time during which an executor may execute a process.

\subsubsection{Workflows}
TODO
\begin{figure}[h]
	\centering
    \includegraphics[scale=0.32]{workflow.png}
	\caption{cron management}
	\label{fig:fig3}
\end{figure}
